Statistical Model Selection for Markov Models of Biomolecular Dynamics

### 2014-04-05 paper/paper2.tex
edit

\begin{document}
\maketitle
\begin{abstract}
Markov state models provide a powerful framework for the analysis of biomolecular conformation dynamics in terms of their metastable states and transition rates. These models provide both a quantitative and comprehensible description of the long-timescale dynamics of large molecular dynamics with a Master equation, and have been successfully used to study protein folding, protein conformational change, and protein-ligand binding. However, to achieve satisfactory performance, existing methodologies often require expert intervention when defining the model's discrete state space. While standard model selection methodologies focus on the minimization of systematic bias and disregard statistical error, we show that by consideration of the states' conditional distribution over conformations, both sources of error can be balanced evenhandedly. Application of these techniques on two 100$\mu s$ molecular dynamics trajectories of the Fip35 WW domain shows agreement with existing techniques based on self-consistency of the model's relaxation timescales, with more suitable results in regimes in which those techniques encourage over-fitting. By removing the need for expert tuning, these methods should reduce modeling bias and lower the barriers to entry in Markvov state model construction.
\end{abstract}

@@

Markov state models (MSMs) are one approach for analyzing MD data sets that are able to smoothly move between high and low-resolution models\cite{Chodera2007Automatic, Prinz2011Markov, Beauchamp2012Simple, Bowman2013Quantitative}. High-resolution  models maintain quantitative agreement with the underlying simulation data, while low-resolution models capture the salient features of the potential energy landscape, sacrificing some degree of model complexity. In an MSM, the dynamics are modeled as a memory-less jump process between a discrete set of conformational states which partition the phase space. Two key quantities which define the MSM are thus the state definitions, an indicator function basis over phase space, and the pairwise transition probabilities or transition rates, which parameterize the kinetics. The matrix of transition probabilities can be used to locate the systems transition paths\cite{Weinan2006Towards}, and its dominant eigenvectors identify the metastable states\cite{Deuflhard2000Identification} and long-timescale dynamical modes\cite{}.

Some explanation of notation...
\begin{align}
s(t) &= \sigma(x(t)), \\
s(t+\tau) &\sim \mathbf{T}_{s(t), k} \sigma_k.
\end{align}

A significant challenge in the automated construction of Markov state models is the choice of the number of states\cite{McGibbon2013Learning}. Although classical Hamiltonian dynamics with $N$ particles form a continuous-time Markov chain in $\mathbb{R}^{6N}$\cite{vanKampen}, the projection of this high-dimensional process into a lower dimensional space does (such as the discrete indicator functions used by MSMs) generally experiences non-Markovian dynamics. In the Morri-Zwanzig approach, this corresponds to the generalized Langevin equation with a nontrivial memory kernel\cite{ZM}. Within the MSM context, on the other hand, this means that first order Markov chains incur a systematic error in their modeling of the high-dimensional system. In particular, when MSM states contain within them kinetic barriers of substantial magnitude, the bias due to the first-order Markov approximation is large. While this source of modeling error can be addressed by increasing the number of states, the reduction in one error comes at the expense of the increase in another. This second source of error is statistical in origin. As the number of states in the model grows, so does the number of parameters required to completely specify the kinetic model between all pairs of states. Because the amount of data is constant, each additional parameter leads to a decrease in the amount of data available per model parameter, which makes the approach susceptible to over-fitting.

@@

\section{Prior Work}
%After diagonalization of the transition matrix, the relaxation of any initial probability distribution can be described by a superposition of exponential relaxation processes, with a characteristic time given by $-\frac{1}{\ln \lambda_i}$.
MSM model selection has traditionally been performed by analyzing the self-consistency of the models with respect to changes in the Markov lag time. As suggested by \citet{Swope2004Describing}, given a Markov model at a lag time of $\tau$, $T(\tau)$, we expect that a model with a lag time of $n\tau$ should merely be the $n^\textnormal{th}$ power of $T(\tau)$:
$$ T(n\tau) = T(\tau)^n $$ As such, the eigenvalues, and correspondingly the relaxation timescales, should remain constant as the lag time increases. The resulting test of a model's quality consists of calculating the eigenspectra of models built at many lag times and choosing models whose eigenvalues have converged.

In practice, while this test has been useful for many \cite{Noe2009Constructing}, it has some flaws. In particular, it is generally observed that most models' timescales do not converge, but steadily increase, making it is difficult to judge whether a particular model passes the test and encouraging the use of ever-more states. Additionally, it is difficult to compare models built on two state decompositions, as neither may have convergent timescales. Often, the choice is made to use the model whose timescales better match an experimental observable.

@@

The appropriate likelihood is instead a path action in phase space, on which the discrete states are merely an indicator function basis. With $s(\mathbf{x})$ as the function mapping conformations into the indicator function basis set, and $\{\mathbf{x}_t\}_{t=1}^N$ an observed trajectory, the likelihood can be written as:

@@

With a discrete, non-overlapping state space, the likelihood of a sampled trajectory can be decomposed into a product of terms of two types: the state to state Markov transition probabilities, $T(s_i \rightarrow s_j) \equiv P(S_{t+1}=s_j | S_{t}=s_i)$, and so-called emission distributions of each state, $P(\mathbf{x} | s(\mathbf{x}))$, the conditional probability of observing a conformation at a given location in phase space given that the conformation, $\mathbf{x}$, is within a certain state, $s_i$. By convention, we define $P(\mathbf{x}| s(\mathbf{x}))$ to be zero for all $\mathbf{x}$ such that $s(\mathbf{x}) \neq s_i$.

This emission distribution can be potentially modeled in multiple ways, each of which, in combination with a transition matrix, parametrizes a different statistical model. \citet{Kellogg2012Evaluation} used a discriminative model with:
\begin{equation}
P(\mathbf{x}_t | s(\mathbf{x}_t)) = \big| \{ \mathbf{x}_k : s(\mathbf{x}_k) = s(\mathbf{x}_t) \} \big|^{-1}
\end{equation} This model has the training data set as its support rather than phase space, which makes it unable to generalize and assign probability to new data. In certain circumstances, this is an undesirable property. For example, protocols that involve fitting and validating models on separate data sets (e.g. cross-validation) are impossible when the statistical model lacks the capacity to describe new data.

Instead, we seek a generative model whose emission distributions are supported on phase space (i.e. standard distributions over $\mathbb{R}^{3N_{atoms}}$). We propose two possible emission distributions, $P(\mathbf{x} | s(\mathbf{x}))$: first, a uniform distribution over the (hyper)volume of each state, as shown in \cref{eq:like_vol}. This model has the advantage of appearing agnostic with respect to intra-state dynamics but is intractable in high dimensions when the states are defined by general polytopes, such as those produced by a data-driven Voronoi tessellation.

%For example, consider two Markov state models sharing the same transition matrix, $T$. In one model, the state emission distributions are highly peaked at specific locations in phase space, whereas in the other model the emission distributions are uniform over the volume of the states. If an observed trajectory does go through the first models' regions of high likelihood, it is appropriately termed a more likely model given the data.

%The simplest definition would be a uniform distribution since we are constructing an MSM that is agnostic to intra-state dynamics. When properly normalized, the uniform distribution takes a value of the reciprocal of the state's volume. This makes for an easy expression, however, volumes in high-dimensional states can be difficult to compute \ref{eq:like_vol}. For this reason, we can also choose other emission distributions that may be more tractable. In fact, \citet{Pelleg2000Xmeans} have used a gaussian emission distribution to define a likelihood to compare two k-means clusterings. The result of this emission distribution, is that the likelihood is proportional to the sum of square distances, which is precisely what is optimized in k-means clustering. If we think of k-centers as a poor man's version of k-means, then we can apply this same likelihood.
\begin{equation}
\label{eq:like_vol}
\begin{split}
%P\left[\{\mathbf{x}_t\}_{t=1}^N\right] d\mathbf{x}^N = \prod_{t=1}^{N-1} T(s(\mathbf{x}_t) \rightarrow s(\mathbf{x}_{t+1})) \\
 P(\mathbf{x}_t | s(\mathbf{x}_t)) = \frac{1}{V_{s(\mathbf{x}_{t})}}
\end{split}
\end{equation}

A more tractable alternative is a Gaussian emission model where $P(\mathbf{x} | s(\mathbf{x}))$ is modeled as multivariate normal \cref{eq:like_mvn}.
\begin{equation}
\label{eq:like_mvn}
\begin{split}
P(\mathbf{x}_t | s(\mathbf{x}_t)) = \Bigg[ \frac{1}{(2 \pi)^\frac{d}{2} |\Sigma_{s(\mathbf{x}_t)}|^\frac{1}{2}} \\
   \exp\left(-\frac{1}{2} (\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)})^T \Sigma_{s(\mathbf{x}_t)}^{-1} (\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)})\right)\Bigg] 
\end{split}
\end{equation} where $\bm{\mu}_{s(\mathbf{x}_t)}$ is the cluster center assigned to $\mathbf{x}_t$ and $\Sigma_{s(\mathbf{x}_t)}$ is the corresponding covariance matrix. For simplicity, we follow \citet{Pelleg2000Xmeans} and employ a spherical Gaussian model with a single shared variance parameter across the states estimated by maximum likelihood, under which it reduces to:
\begin{equation}
\label{eq:like_mvn2}
 P(\mathbf{x}_t | s(\mathbf{x}_t)) = \frac{1}{\left(2 \pi \hat{\sigma}^2\right)^{\frac{d}{2}}} \exp\left(-\frac{||\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)}||^2}{2 \hat{\sigma}^2} \right)
\end{equation} where
\begin{equation}
\label{eq:mle_sigma}
\hat{\sigma}^2 = \frac{1}{N - k} \sum_{t=1}^N || \mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)} ||^2
\end{equation} and $k$ is the number of states.

@@

Instead, we explore three alternative procedures for choosing the number of states in an MSM: cross validation, Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} and the Akaike information criterion (AIC)\cite{Akaike1974AIC}. Cross-validation attempts to directly measure the Markov model's generalization error. First, the model is parameterized by building both the state space and transition matrix on a subset of the available molecular dynamics trajectories, then the likelihood is evaluated on the left-out portion. This scheme can be repeated many times with different partitions of the data set. The AIC and BIC are augmented likelihood function approaches which do not require leaving out portions of the available data during fitting, and instead employ asymptotic approximations to the Bayesian evidence to directly penalize models with additional free parameters.
\begin{equation}
\label{eq:bic}
\mathrm{BIC} \equiv -2\cdot \ln L + \kappa \cdot \ln N
\end{equation}
\begin{equation}
\label{eq:aic}
\mathrm{AIC} \equiv -2\cdot \ln L + 2 \cdot \kappa
\end{equation} where $L$ is the likelihood, $\kappa$ is the number of free parameters, and $N$ is the number of data points, assumed to be independent and identically distributed. Model selection is performed by a minimization of the criterion.

@@

\begin{figure}
\centering
\includegraphics[width=3in]{figs_final/mull_ww.png}
\caption{Systems studied in this work. (A) Langevin dynamics on the two dimensional M\"{u}ller potential \cite{Muller1980Reaction}. (b) 200 $\mu s$ of dynamics of the Fip35 WW domain\cite{Liu2008Experimental}, courtesy of D.E. Shaw research \cite{Shaw2010Atomic}.}
\end{figure}

@@

For cross validation, we find it essential to use a dense prior during transition matrix estimation. Because the maximum likelihood transition matrix fit on a subset of the data assigns probability zero to unobserved transitions, we observe cross validation log likelihoods of negative infinity in all but the most data-rich regimes. We used a Dirichlet prior with a pseudo-count of $\frac{1}{k}$ for each possible transition, where $k$ is the number of states in the model.

\section{Results and Discussion}
\subsection{M\"{u}ller Potential}
We simulated two trajectories with $10^6$ steps of Langevin dynamics on the M\"{u}ller potential\cite{Muller1980Reaction}. Simulations were performed with a timestep $t=0.1$, friction coefficient of $\gamma=10^3$, and $kT=15$. The first trajectory was clustered using the $k$-centers clustering algorithm, and state volumes were computed for the uniform emission model using $10^5$ rounds of Monte Carlo rejection sampling with a cutoff of $R=0.28$ from the 50 state model's centroids. The second trajectory was used as a test set. All MSMs were built using a lag time of $30 steps$.

The volumes did not change drastically upon doing ten times more Monte Carlo rejection sampling, and the likelihood remained essentially constant, which indicates that the use of approximation volumes is sufficient to obtain robust estimates of the likelihood.

\begin{figure}[h]
\centering
\includegraphics[width=3in]{figs_final/mull_likes.png}
\caption{We calculated the training and test set log-likelihoods and model selection criteria for models built with 50-1000 states using both the uniform ({\bf A}) and multivariate normal ({\bf B}) emission distributions. Dashed lines represent the optimal model as defined by the AIC, BIC, or test set log-likelihood and form the boundaries of the optimal window (shaded dark gray). For both likelihoods, the BIC penalizes complexity more than the AIC and test set log-likelihood. The optimal window is consistent with the convergence of the relaxation timescales (right frames).}
\label{fig:mullerlike}
\end{figure}

As shown in \cref{fig:mullerlike}, models built with too few states achieve a drastically reduced likelihood, but above a threshold region the likelihood increases relatively slowly. The penalty on the number of parameters in \cref{eq:bic} and \cref{eq:aic} begins to dominate. The optimal models, according to the three methods are between 200 and 600 states for this system, which is consistent with the convergence of the relaxation timescales of the models.

The AIC and BIC penalize the larger state models much more than the test set log-likelihood. On the two-dimensional potential, in our data-rich regime it is difficult to produce an over-fit model. Cross-validation requires fitting multiple models on subsets of the data, and so is not feasible for larger systems in the data-poor regime.

We also tested the Gaussian likelihood on MSMs built using k-means clustering on the M\"uller potential. The training and test set log-likelihoods, BIC, and AIC are all optimized by a five state model (\cref{fig:kmeans_mull}). The decrease in the training set log-likelihood can be explained by a trade-off inherent in \cref{eq:like}, where adding more states leads to an improvement in the emission distribution term but causes the transition matrix term to suffer. This trade-off illustrates a deficiency in the step-wise approach in which neither the transition matrix nor the state space is parameterized to explicitly optimize the full \cref{eq:like}.

\begin{figure}
\centering
\includegraphics[width=3in]{figs_final/kmeans_like2_lbl.png}
\caption{MSMs were built using k-means clustering on the M\"uller potential simulations. Since k-means optimizes an objective function that is related to the emission probability, our likelihood picks only a few states as the most likely model. The likelihood decreases as new states are introduced, due to a competition between the transition matrix term and emission distribution term in \cref{eq:like}.}
\label{fig:kmeans_mull}
\end{figure}

@@

We reanalyzed two ultra-long 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain\cite{Liu2008Experimental}, provided courtesy of D.E. Shaw Research \cite{Shaw2010Atomic}. Because the likelihood calculations involving the uniform distribution emission model have exponential complexity in the dimensionality of the state space we first preprocess the trajectories with time-structure based independent components analysis (tICA) \cite{Schwantes2013Improvements, Perez2013Identification}, extracting only the four slowest uncorrelated linear combinations of residue-residue distances. All MSMs were built using a lag time of $50 ns$.

The complexity of evaluating the Gaussian emission model likelihood does not increase with respect to dimension. Thus we used the Gaussian emission model with both the four-dimensional tICA models and those built using k-centers clustering with the minimal cartesian root-mean squared deviation (RMSD) over Fip35's 252 non-redundant heavy atoms, computed using the quaternion characteristic polynomial method\cite{Theobald2005Rapid}. For evaluating the Gaussian emission likelihood, we used a value of $3 \cdot 252 - 6 = 750$ for the dimension of the emission distribution.

\begin{figure}
\centering
\includegraphics[width=3in]{figs_final/ww_likes.png}
\caption{On models whose states were defined after projecting into the four-dimensional tICA subspace, we computed likelihoods based on the uniform ({\bf A}) and Gaussian ({\bf B}) emission models. Models built using the RMSD metric were evaluated with the Gaussian emission model ({\bf C}). Dashed lines represent the optimal model as defined by the AIC, BIC, or test set log-likelihood and form the boundaries of the optimal window (shaded dark gray). Test set log-likelihoods are omitted in {\bf C} as the dense matrix algebra is intractable for $k > 10^3$.\label{fig:ww}}
\end{figure}

@@

The two 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain simulated by \citet{Shaw2010Atomic} have been used to parameterize multiple Markov state models. \citet{Lane2011Markov} proposed a 26,000 state MSM after clustering the data by RMSD, whereas Kellogg et al.,\cite{Kellogg2012Evaluation} clustering on contact maps or secondary structure, proposed 100 and 175 state models, respectively. Consistent with this spread, we find the optimal number of states to be highly dependent on the space in which the data is clustered. With RMSD, the Gaussian likelihood indicates that Markov models built with on the order of 5,000 states optimally balance model quality and generalization error. On the other hand, the tICA method's identification of slow order parameters in the simulations has the effect of reducing the number of states necessary when constructing Markov models. Even with 200-600 states, Markov models constructed using tICA are able to accurately identify the simulations folding process on the 1-10 $\mu s$ timescale, as well as the faster near-native dynamics identified in this data set by McGibbon and Pande\cite{McGibbon2013Learning}.

@@

\bibliography{bibliography}
\end{document}

>>>>

\begin{document}
\maketitle
\begin{abstract}
Markov state models provide a powerful framework for the analysis of biomolecular conformation dynamics in terms of their metastable states and transition rates. These models provide both a quantitative and comprehensible description of the long-timescale dynamics of large molecular dynamics with a Master equation, and have been successfully used to study protein folding, protein conformational change, and protein-ligand binding. However, to achieve satisfactory performance, existing methodologies often require expert intervention when defining the model's discrete state space. While standard model selection methodologies focus on the minimization of systematic bias and disregard statistical error, we show that by consideration of the states' conditional distribution over conformations, both sources of error can be balanced evenhandedly. Application of techniques that consider both systematic bias and statistical error on two 100$\mu s$ molecular dynamics trajectories of the Fip35 WW domain shows agreement with existing techniques based on self-consistency of the model's relaxation timescales, with more suitable results in regimes in which those timescale-based techniques encourage over-fitting. By removing the need for expert tuning, these methods should reduce modeling bias and lower the barriers to entry in Markov state model construction.
\end{abstract}

@@

Markov state models (MSMs) are one approach for analyzing MD data sets that are able to smoothly move between high and low-resolution models\cite{Chodera2007Automatic, Prinz2011Markov, Beauchamp2012Simple, Bowman2013Quantitative}. High-resolution  models maintain quantitative agreement with the underlying simulation data, while low-resolution models capture the salient features of the potential energy landscape, sacrificing some degree of model complexity. In an MSM, the dynamics are modeled as a memory-less jump process between a discrete set of conformational states which partition the phase space. Two key quantities which define the MSM are thus the state definitions, an indicator function basis over phase space, and the pairwise transition probabilities or transition rates, which parameterize the kinetics. The matrix of transition probabilities can be used to locate the systems transition paths\cite{Weinan2006Towards}, and its dominant eigenvectors identify the metastable states\cite{Deuflhard2000Identification} and long-timescale dynamical modes\cite{who?}.

%The trajectory in phase space ($\mathbf{x}_t$) is assigned to one of $k$ states. We represent this assignment by the $s(\cdot)$ function. The dynamics in state space are described by the transition probability matrix. Since the model is Markovian, the probability that a given trajectory in state $s(\mathbf{x}_t)$ will be in some state $i$ is given by $\mathbf{T}_{s(\mathbf{x}_t), i}$:
%\begin{equation}
%\label{eq:markov_state}
%s(\mathbf{x}_{t+\tau}) \sim_d \mathbf{T}_{s(\mathbf{x}_t), i}
%\end{equation}

A significant challenge in the automated construction of Markov state models is the choice of the number of states\cite{McGibbon2013Learning}. Although classical Hamiltonian dynamics with $N$ particles form a continuous-time Markov chain in $\mathbb{R}^{6N}$\cite{vanKampen}, the projection of this high-dimensional process into a lower dimensional space (such as the discrete indicator functions used by MSMs) is generally governed by non-Markovian dynamics. \hl{In the Morri-Zwanzig approach, this corresponds to the generalized Langevin equation with a nontrivial memory kernel}(What is this about?)\cite{ZM}. Within the MSM context, on the other hand, this means that first order Markov chains incur a systematic error in their modeling of the high-dimensional system. In particular, when MSM states contain within them kinetic barriers of substantial magnitude, the bias due to the first-order Markov approximation is large. While this source of modeling error can be addressed by increasing the number of states, the reduction in one error comes at the expense of the increase in another. This second source of error is statistical in origin. As the number of states in the model grows, so does the number of parameters required to completely specify the kinetic model between all pairs of states. Because the amount of data is constant, each additional parameter leads to a decrease in the amount of data available per model parameter, which makes the approach susceptible to over-fitting.

@@

\section{Background}
\begin{itemize}
\item For some system, we are interested in describing the arbitrarily long timescale dynamics. 
\item define the phase space, and then an ensemble of systems propagating through phase space. 
\item There exists an operator such that operator of p(0) is p(t). Formally this operator is dependent on the hamiltonian
\item This operator has a set of characteristic timescales (and eigenvectors) such that p at t + ntau can be decomposed into a sum of exponential relaxation processes along each eigenvector (show the equation)
\item A central focus for MD, is the estimation of the transfer operator 
\item Calculating the hamiltonian integral is not possible, but estimators are available based on sampling the hamiltonian (MD) 
\item One framework for estimating the transfer operator is the discrete-time Markov State Model (MSM), which is a projection of the operator onto an indicator function basis set. 
\item The MSM is parameterized by a set of states corresponding to the indicator functions defined on phase space as well as the transition probabilities for transitioning between any state to any other state at a lag time (tau)
\item It has been shown that using a finite number of discrete states in phase space, estimates of the transfer operator's timescales are biased to be too fast. 
\item This result implies that one can always do better by varying the parameters  to increase the timescales (include the idea of a variational principle)
\item However, these results are proved based on the assumption of infinite data, which is never the case
\end{itemize}
\section{another section}
\begin{itemize}
\item We always have finite data, and so this variational approach ignores another source of error due to the statistical error
\item What we really want is a model that reproduces the dynamics due to the hamiltonian
\item In practice, the transfer operator is inaccessible, and what you have is data and a model that attempts to describe the data.
\item But the only way to access the true transfer operator is to sample from it and produce new data. Therefore, we should actually attempt to estimate a model that can best predict new data.
\item This MSM has since been referred to as an object that takes in data, but in fact (as it is generative) we are parameterizing it to generate new data. 
\item So if we build many MSMs, the ``best" model is the one that is most likely to have produced the newly observed data. 
\item This means that we should construct a model such that it assigns a high probability to the new data. 
\item This is precisely the idea of cross-validation or BIC/AIC (maybe we should define them here).
\item The true data is a random variable (that we can sample), but in an MSM, the trajectory sampled is a random variable
\end{itemize}

\section{bayesian statistics}
\begin{itemize}
\item in the bayesian world, we would integrate over all possible models, but if we want a single model, then we should pick the most likely one
\end{itemize}

Given a Markov model parameterized at a lag time of $\tau$, we can propagate ensembles of trajectories via the transition probability matrix:
$$\mathbf{p}_\tau^T = \mathbf{p}_0^T T(\tau)$$ where $\mathbf{p}_t$ is a probability distribution over states at time $t$ and $T(\tau)$ is a matrix with entries corresponding to the probability of transferring from one state to another at a time of $\tau$. As suggested by \citet{Swope2004Describing}, given a Markov model at a lag time of $\tau$, $T(\tau)$, we expect that a different model with a lag time of $n\tau$ should merely be the $n^\textnormal{th}$ power of $T(\tau)$:
$$\mathbf{p}_{n\tau}^T = \mathbf{p}_0^T T(n\tau) = \mathbf{p}_0^T \left[T(\tau)\right]^n $$ As such, the eigenvalues and eigenvectors of models built at many lag times are related. Specifically, when building multiple models for the same dataset, the eigenvectors should be invariant to changing lag time, while the eigenvalues change via the relation:
$$\lambda(\tau)^n = \lambda(n \tau)$$ The largest, non-unit eigenvalues correspond to the slowest timescales in the data. It's often useful to compute the slowest implied timescales ($t_i$) via the relation:
$$t_i = - \frac{\tau}{\log \lambda_i} $$ These implied timescales should be invariant with respect to the lag time and therefore provide a useful tool for testing the quality of an MSM. Specifically, many models are built at varying lag times, and one is selected by choosing the model whose timescales have converged.

\section{Prior Work}
%After diagonalization of the transition matrix, the relaxation of any initial probability distribution can be described by a superposition of exponential relaxation processes, with a characteristic time given by $-\frac{1}{\ln \lambda_i}$.
MSM model selection has traditionally been performed by analyzing the self-consistency of the models with respect to changes in the Markov lag time. In practice, while this test has been useful for many \cite{Noe2009Constructing}, it has some flaws. In particular, it is generally observed that most models' timescales do not converge, but steadily increase, making it difficult to judge whether a particular model passes the test and encouraging the use of ever-more states. Additionally, it is difficult to compare models built on two state decompositions, as neither may have convergent timescales. Often, the choice is made to use the model whose timescales better match an experimental observable.

@@

The appropriate likelihood is instead a path action in phase space, on which the discrete states are merely an indicator function basis. With $s(\mathbf{x})$ as the function that maps a conformation in phase space ($\mathbf{x}$) to the index of a state in the discrete state space, and $\{\mathbf{x}_t\}_{t=1}^N$ an observed trajectory, the likelihood can be written as:

@@

With a discrete, non-overlapping state space, the likelihood of a sampled trajectory can be decomposed into a product of terms of two types: the state to state Markov transition probabilities, $T(s_i \rightarrow s_j) \equiv P(S_{t+1}=s_j | S_{t}=s_i)$, and so-called emission distributions of each state, $P(\mathbf{x} | s(\mathbf{x}))$, the conditional probability of observing a conformation at a given location in phase space given that the conformation, $\mathbf{x}$, is within a certain state, $s(\mathbf{x})$. As we are using  non-overlapping states, the conditional probability of observing a conformation, $\mathbf{x}$ from state $i$ is zero unless $\mathbf{x}$ is assigned to $i$. Mathematically, this means:
$$ P(\mathbf{x} | i) = 0 \;\;\;\; \forall \mathbf{x} \textnormal{ s.t. } s(\mathbf{x}) \neq i$$

This emission distribution can be modeled in multiple ways, each of which, in combination with a transition matrix, parametrizes a different statistical model. \citet{Kellogg2012Evaluation} used a discriminative model with:
\begin{equation}
P(\mathbf{x}_t | s(\mathbf{x}_t)) = \big| \{ \mathbf{x}_k : s(\mathbf{x}_k) = s(\mathbf{x}_t) \} \big|^{-1}
\end{equation} where $| \cdot |$ denotes the cardinality of a set. This is a discriminative model, whose support is the training data set rather than phase space, which makes it unable to generalize and assign probability to new data. In certain circumstances, this is an undesirable property. For example, protocols that involve fitting and validating models on separate data sets (e.g. cross-validation) are impossible when the statistical model lacks the capacity to describe new data.

Instead, we seek a generative model whose emission distributions are supported on phase space (i.e. standard distributions over $\mathbb{R}^{3N_{atoms}}$). We propose one possible emission distributions, $P(\mathbf{x} | s(\mathbf{x}))$: a uniform distribution over the (hyper)volume of each state, as shown in \cref{eq:like_vol}. 
\begin{equation}
\label{eq:like_vol}
\begin{split}
%P\left[\{\mathbf{x}_t\}_{t=1}^N\right] d\mathbf{x}^N = \prod_{t=1}^{N-1} T(s(\mathbf{x}_t) \rightarrow s(\mathbf{x}_{t+1})) \\
 P(\mathbf{x}_t | s(\mathbf{x}_t)) = \frac{1}{V_{s(\mathbf{x}_{t})}}
\end{split}
\end{equation} 
This model has the advantage of appearing agnostic with respect to intra-state dynamics but is intractable in high dimensions when the states are defined by general polytopes, such as those produced by a data-driven Voronoi tessellation. 
%For example, consider two Markov state models sharing the same transition matrix, $T$. In one model, the state emission distributions are highly peaked at specific locations in phase space, whereas in the other model the emission distributions are uniform over the volume of the states. If an observed trajectory does go through the first models' regions of high likelihood, it is appropriately termed a more likely model given the data.

%The simplest definition would be a uniform distribution since we are constructing an MSM that is agnostic to intra-state dynamics. When properly normalized, the uniform distribution takes a value of the reciprocal of the state's volume. This makes for an easy expression, however, volumes in high-dimensional states can be difficult to compute \ref{eq:like_vol}. For this reason, we can also choose other emission distributions that may be more tractable. In fact, \citet{Pelleg2000Xmeans} have used a gaussian emission distribution to define a likelihood to compare two k-means clusterings. The result of this emission distribution, is that the likelihood is proportional to the sum of square distances, which is precisely what is optimized in k-means clustering. If we think of k-centers as a poor man's version of k-means, then we can apply this same likelihood.

@@

Instead, we explore three alternative procedures for choosing the number of states in an MSM: cross validation, Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} and the Akaike information criterion (AIC)\cite{Akaike1974AIC}. Cross-validation attempts to directly measure the Markov model's generalization error. First, the model is parameterized by building both the state space and transition matrix on a subset of the available molecular dynamics trajectories, then the likelihood is evaluated on the left-out portion. This scheme can be repeated many times with different partitions of the data set. The AIC and BIC are augmented likelihood function approaches which do not require leaving out portions of the available data during fitting, and instead employ asymptotic approximations to the Bayesian evidence to directly penalize models with additional free parameters.
\begin{equation}
\label{eq:bic}
\mathrm{BIC} \equiv -2\cdot \ln L + \left(\ln N \right)\cdot \kappa
\end{equation}
\begin{equation}
\label{eq:aic}
\mathrm{AIC} \equiv -2\cdot \ln L + \left(2\right) \cdot \kappa
\end{equation} where $L$ is the likelihood, $\kappa$ is the number of free parameters, and $N$ is the number of data points, assumed to be independent and identically distributed. Model selection is performed by a minimization of the criterion.

@@

\begin{figure}
\centering
\includegraphics[width=3in]{figs_final/mull_ww.png}
\caption{Systems studied in this work. (A) Langevin dynamics on the two dimensional M\"{u}ller potential \cite{Muller1980Reaction}. (b) 200 $\mu s$ of dynamics of the Fip35 WW domain\cite{Liu2008Experimental}, courtesy of D.E. Shaw research \cite{Shaw2010Atomic}.}
\label{fig:pics}
\end{figure}

@@

Because the maximum likelihood transition matrix fit on a subset of the data assigns probability zero to unobserved transitions, we observe cross validation log likelihoods of negative infinity in all but the most data-rich regimes. For this reason, we find it essential to use a dense prior during transition matrix estimation. Using a prior ensures that all transitions have a nonzero (albeit small) probability. We used a Dirichlet prior with a pseudo-count of $\frac{1}{k}$ for each possible transition, where $k$ is the number of states in the model\cite{msmb2?, nina?, wherever this happens to be}.

\section{Results and Discussion}
\subsection{M\"{u}ller Potential}
The M\"{u}ller potential, is a simple two-dimensional surface with three energy minima. We simulated two trajectories with $10^6$ steps of Langevin dynamics on the M\"{u}ller potential\cite{Muller1980Reaction}. Simulations were performed with a timestep $t=0.1$, friction coefficient of $\gamma=10^3$, and $kT=15$.

The first trajectory was clustered using the $k$-centers clustering algorithm, and state volumes were computed for the uniform emission model using $10^5$ rounds of Monte Carlo rejection sampling with a cutoff of $R=0.28$ from the 50 state model's centroids. (To give a sense of scale, the x-axis of the M\"{u}ller potential shown in \cref{fig:pics} is 4 units across). The second trajectory was used as a test set. All MSMs were built using a lag time of 30 steps.

The volumes did not change drastically upon doing ten times more Monte Carlo rejection sampling, and the likelihood remained essentially constant, which indicates that the use of approximate volumes is sufficient to obtain robust estimates of the likelihood.

\begin{figure}[h]
\centering
\includegraphics[width=4in]{figs_final/mull_likes.png}
\caption{\hl{Fix this figure so that only part (A) is depicted.} We calculated the training and test set log-likelihoods and model selection criteria for models built with 50-1000 states using the uniform emission distribution. Dashed lines represent the optimal model as defined by the AIC, BIC, or test set log-likelihood and form the boundaries of the optimal window (shaded dark gray). For both likelihoods, the BIC penalizes complexity more than the AIC and test set log-likelihood. The optimal window is consistent with the convergence of the relaxation timescales (right frames).}
\label{fig:mullerlike}
\end{figure}

As shown in \cref{fig:mullerlike}, models built with too few states achieve a drastically reduced likelihood, but above a threshold region the likelihood increases relatively slowly. The penalty on the number of parameters in \cref{eq:bic} and \cref{eq:aic} begins to dominate. The optimal models, according to the BIC, AIC, and test log-likelihood are between 200 and 600 states for this system, which is consistent with the convergence of the relaxation timescales of the models.

The AIC and BIC penalize the larger state models much more than the test set log-likelihood. It's not clear why the test log-likelihood is more lenient, however, it suggests that when possible, one should use a test log-likelihood approach as opposed to an approximate method like the AIC or BIC. This cross-validation requires fitting multiple models on subsets of the data, and so is not feasible for larger systems in the data-poor regime.

@@

We reanalyzed two ultra-long 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain\cite{Liu2008Experimental}, provided courtesy of D.E. Shaw Research \cite{Shaw2010Atomic}. Because the likelihood calculations involving the uniform distribution emission model have exponential complexity in the dimensionality of the state space we first preprocess the trajectories with time-structure based independent components analysis (tICA) \cite{Schwantes2013Improvements, Perez2013Identification}, extracting only the four slowest uncorrelated linear combinations of residue-residue distances. All MSMs were built using a lag time of 50 ns.

\begin{figure}
\centering
\includegraphics[width=3in]{figs_final/ww_likes.png}
\caption{\hl{Fix this figure so that only part (A) is depicted.} On models whose states were defined after projecting into the four-dimensional tICA subspace, we computed likelihoods based on the uniform emission models. Dashed lines represent the optimal model as defined by the AIC, BIC, or test set log-likelihood and form the boundaries of the optimal window (shaded dark gray). \label{fig:ww}}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=6in]{figs_final/timescales_suck2.pdf}
\caption{For MSMs of the WW domain built on the top three PCA components,} 
\end{figure}

%The two 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain simulated by \citet{Shaw2010Atomic} have been used to parameterize multiple Markov state models. \citet{Lane2011Markov} proposed a 26,000 state MSM after clustering the data by RMSD, whereas Kellogg et al.,\cite{Kellogg2012Evaluation} clustering on contact maps or secondary structure, proposed 100 and 175 state models, respectively. Consistent with this spread, we find the optimal number of states to be highly dependent on the space in which the data is clustered. With RMSD, the Gaussian likelihood indicates that Markov models built with on the order of 5,000 states optimally balance model quality and generalization error. On the other hand, the tICA method's identification of slow order parameters in the simulations has the effect of reducing the number of states necessary when constructing Markov models. While only using 200-600 states, Markov models constructed using tICA were still able to recapitulate the folding timescale as well as the faster processes identified by McGibbon and Pande\cite{McGibbon2013Learning}.

@@

\bibliography{bibliography}

\section{Appendix}
\subsection{Other emission distributions}
